Special Constructions in Categories

### 2017-08-26 papers/talk-6/paper-6.tex
writeup6: Zweite Korrektur-Iteration

  Let $A$ be another set and $f_1 \from A \to X, f_2 \from A \to Y$ functions.
  If we want to define a function $\fbar \from A \to X \times Y$, such that
  $p_1(\fbar(a)) = f_1(a)$ and $p_2(\fbar(a)) = f_2(a)$ for every $a \in A$ holds,
  there is only one possible way to define $\fbar \from A \to X \times Y$:
  $$\fbar(a) := \pair{f_1(a)}{f_2(a)}.$$
  By construction we have $p_1(\fbar(a)) = p_1(f_1(a), f_2(b)) = f_1(a)$ and $p_2(\fbar(a)) = p_2(f_1(a), f_2(a)) = f_2(a)$.
  If we have another $\ftilde \from A \to X \times Y$ with $p_1(\ftilde(a)) = f_1(a)$ and $p_2(\ftilde(a)) = f_2(a)$ we show $\fbar(a) = \ftilde(a)$ for every $a \in A$:
  $$\fbar(a) = \pair{f_1(a)}{f_2(a)} = \pair{p_1(\ftilde(a)}{p_2(\ftilde(a))} = \ftilde(a).$$
  The last step is because we have $\pair{p_1(z)}{p_2(z)} = z$ for every $z \in X \times Y$.
\end{example}

@@

\begin{proof}
  Let $X, Y \in \cat{A}$ and $P$ be a product of $X$ and $Y$ together with $p_1 \from P \to X, p_2 \from P \to Y$
  and $P'$ be a product of $X$ and $Y$ together with $p'_1 \from P' \to X, p'_2 \from P' \to Y$.
  We apply definition \ref{def:prod} multiple times and get the following commuting diagram:
  \[ \xymatrix{
    & P \ar@/_/[ldd]_{p_1} \ar@/^/[rdd]^{p_2} & \\
    & P \ar[ld]^{p_1} \ar@/_/@{.>}[dd]|{\fbar} \ar@{.>}[u]|{f} \ar[rd]_{p_2} & \\
    X & & Y \\
    & P' \ar[lu]_{p_1'} \ar@/_/@{.>}[uu]|{\ftilde} \ar@{.>}[d]|{f'} \ar[ru]^{p'_2} & \\
    & P' \ar@/^/[luu]^{p'_1} \ar@/_/[ruu]_{p'_2} &
  } \]
  for exactly one $f, f', \fbar, \ftilde$. With the inner triangles we get:
  \begin{alignat*}{2}
    p_1  \of \ftilde &= p'_1 &\qquad p_2  \of \ftilde &= p'_2 \\
    p'_1 \of \fbar   &= p_1  &       p'_2 \of \fbar   &= p_2
  \end{alignat*}
  By substitution we get:
  \begin{alignat*}{2}
    p_1   \of (\ftilde \of \fbar) &= p_1  &\qquad p_2  \of (\ftilde \of \fbar) &= p_2 \\
    p'_1  \of (\fbar \of \ftilde) &= p'_1 &       p'_2 \of (\fbar \of \ftilde) &= p'_2
  \end{alignat*}
  With the outer triangles we get:
  \begin{alignat*}{5}
    p_1  &\;\of&&\; f  &= p_1  &\qquad p_2  &\;\of&\; f  &= p_2 \\
    p'_1 &\;\of&&\; f' &= p'_1 &\qquad p'_2 &\;\of&\; f' &= p'_2
  \end{alignat*}
  Note that $f$ and $f'$ are \emph{unique} and the first equations hold for $f = 1_P$ and $f = \ftilde \of \fbar$, thus $\ftilde \of \fbar = 1_P$.
  The second equations holds for $f' = 1_{P'}$ and for $f' = \fbar \of \ftilde$, thus $\fbar \of \ftilde = 1_{P'}$.
  Thus $P$ and $P'$ are isomorphic. \qedhere
\end{proof}

@@

\begin{example}[Posets]
  \label{ex:prod:poset}
  Let $\pair{A}{\le}$ be a poset.
  A \demph{lower bound} for two objects $x, y \in A$ is a element $a \in A$ such that $a \le x$ and $a \le y$.
  A \demph{greatest lover bound} of $x, y \in A$ is a lower bound $z \in A$ of $x$ and $y$ such that for all lower bounds $a \in A$ of $x$ and $y$, $a \le z$.
  With other words: $z$ is a greatest lower bound of $x$ and $y$ if and only if
  $$z \le x \land z \le y \land \forall a \in A, a \le x \land a \le y \Rightarrow a \le z.$$

  When we see the poset $\pair{A}{\le}$ as a category $\catA$ and have a greatest lower bound $z$ of $x, y \in A$, then $z$ is the product of $x$ and $y$.
  To see this, we apply the definition \ref{def:prod}.
  Let $x, y \in \catA$ and $z \in \catA$ be a greatest lower bound of $x$ and $y$.
  Let $a \in \catA$ be another object.
  If we have $f_1 \from a \to x$ and $f_2 \from a \to y$ then by definition of $\catA$, $a$ is a lower bound of $x$ and $y$.
  Now because $z$ is a greatest lower bound, $a \le z$ holds and therefore is a unique map $\fbar \from a \to z$.
  The diagram commutes trivially.
\end{example}

\begin{example}[Direct sum of vector spaces]
  \label{ex:prod:vspace}
  Let $K$ be a field with the operations $+_K$ and $\cdot_K$ and $V, W$ vector spaces over this field with the operations $+_V, \cdot_V, +_W, \cdot_W$.
  Then we define\footnote{One should verify that the defined structure is indeed a vector field.}
  the \demph{direct sum} $V \oplus W := \bigl(V \times W, +, \cdot \bigr)$ as a new vector field with the following operations:
  \begin{alignat*}{4}
    \pair{v_1}{w_1} + \pair{v_2}{w_2} &:= \pair{v_1 +_V v_2}{w_1 +_W w_2}           &\quad& \forall v_1, v_2 \in V, \forall w_1, w_2 \in W \\
    \alpha \cdot \pair{v}{w}          &:= \pair{\alpha \cdot_V v}{\alpha \cdot_W w} &     & \forall v \in V, \forall w \in W
  \end{alignat*}
  We can show that $U \oplus V$ is the product of $U$ and $V$ in $\Vect$.
  The proof follows the same structure as in example \ref{ex:prod:cart}.
\end{example}

@@

\begin{example}[Direct sum of vector spaces]
  \label{ex:coprod:vspace}
  In the example \ref{ex:prod:vspace} we have seen that the direct sum of two vector spaces $U$ and $V$ is the product of $V$ and $W$ in $\Vect$.
  We will now see that it is \emph{also} the coproduct.
  In general products and coproducts are not the same as we have seen in the examples \ref{ex:prod:cart} and \ref{ex:coprod:dunion}.
  First we have to define the two coprojection functions $p_1 \from V \to V \oplus W, p_2 \from W \to V \oplus W$:
  \begin{alignat*}{2}
    p_1(x) &:= \pair{x}{0} \\
    p_2(y) &:= \pair{0}{y}
  \end{alignat*}
  Then let $A \in \Vect$ and $f_1 \from V \to A, f_2 \from W \to A$ be vector morphisms.
  We define the vector morphism $\fbar \from V \oplus W \to A$ with $\fbar(v, w) := f_1(v) + f_2(w)$.
  By construction and because $f_1$ and $f_2$ are vector morphism we have:
  \begin{alignat*}{4}
    \fbar(p_1(u)) &= \fbar(u, 0) &= f_1(u) + f_2(0) &= f_1(u) \\
    \fbar(p_2(v)) &= \fbar(0, v) &= f_1(0) + f_2(v) &= f_2(v).
  \end{alignat*}
  Now we have to show that $\fbar$ is the only vector morphism with this property.
  Let $\ftilde \from V \oplus W \to A$ be another vector morphism with $\ftilde(p_1(u)) = f_1(u)$ and $\ftilde(p_2(v)) = f_2(v)$.
  Then we have for every $v \in V$ and $w \in W$:
  $$ \fbar(v,w) = f_1(v) + f_2(w)\ = \ftilde(p_1(v)) + \ftilde(p_2(w)) = \ftilde(p_1(v) + p_2(w)) = \ftilde(v, w). $$
  Thus $\fbar = \ftilde$ and $V \oplus W$ is indeed the product of $V$ and $W$ in the category $\Vect$.
\end{example}

\section*{Equalizers and Coequalizer}

@@

\begin{definition}[Coequalizer]
  \label{def:coequa}
  Let $\cat{A}$ be a category and
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y
  } $
  a diagram in $\cat{A}$.
  A \demph{coequalizer} of this diagram is a commuting diagram called \demph{cofork} \\
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^i & C
  } $
  with the property that for all other coforks
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^f & A
  } $
  in $\cat{A}$, there exists a unique map $\fbar\from C \to A$ such that
  \[ \xymatrix{
    & & A  \\
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[ru]^f \ar[rd]_i \\
    & & C \ar@{.>}[uu]|{\fbar}
  } \]
  commutes.
\end{definition}

@@

\begin{definition}[Symmetric reflexive transitive closure]
  Let $R$ be a binary relation over a set $X$.
  $\hat{R} := R \cup R^{-1}$ is the symmetric closure of $R$ where $R^{-1}$ is defined by
  $R^{-1} := \setMap{\pair{b}{a}}{\pair{a}{b} \in R}$.
  The \demph{symmetric transitive closure} of $R$ is $\sim := \bigcup_{n\in\nat} {\hat R}^n$
  where ${\hat R}^n$ is defined recursively with
  ${\hat R}^0 := \setMap{\pair{a}{a}}{a \in X}$ and
  ${\hat R}^{n+1} := R_n \of \hat R = \setMap{\pair{a}{c}}{\pair{a}{b} \in {\hat R}^n \land \pair{b}{c} \in \hat R}$.
  It can be shown that $\sim$ is the \emph{least equivalence relation} containing $R$.
\end{definition}

@@

  To define the function $\fbar \from C \to A$ we need a choice function $\choice \cdot \from C \to Y$ that returns for every equivalence class in $\sim$ a member of it, i.~e.
  $\forall c \in C, \forall y \in Y, y \in c \Leftrightarrow y \sim \choice{c}$
  which is equivalent to $\forall c \in C, \class {\choice c} = c$.

@@

\begin{definition}[Pullback]
  \label{def:pullback}
  Let $\cat{A}$ be a category and
  $ \xymatrix{
    X \ar[r]^s & Z & Y\ar[l]_t \\
  } $
  a diagram in $\cat{A}$.
  A \demph{pullback} of this diagram is a \demph{pullback square}
  \[ \xymatrix{
    P \ar[r]_{p_2} \ar[d]^{p_1} & Y \ar[d]_{t} \\
    X \ar[r]^s & Z
  } \]
  with the property that for all other pull back squares
  \[ \xymatrix{
    A \ar[r]_{f_2} \ar[d]^{f_1} & Y \ar[d]_{t} \\
    X \ar[r]^s & Z
  } \]
  in $\cat{A}$, there exists a unique map $\fbar\from A \to P$ such that
  \[ \xymatrix{
    A \ar[drr]^{f_2} \ar[ddr]_{f_1} \ar@{.>}[dr]|{\fbar} & & \\
    & P \ar[r]_{p_2} \ar[d]^{p_1} & Y \ar[d]_{t} \\
    & X \ar[r]^s & Z
  } \]
  commutes.
\end{definition}

The pullback is also called ``fibred product''. This will be clear in the following example:
\todo{Besser erklären, was das Beispiel damit zu tun hat}

\begin{example}[Fibred product]
  When the pullback square
  \[ \xymatrix{
    P \ar[r]_{p_2} \ar[d]^{p_1} & Y \ar[d]_{t} \\
    X \ar[r]^s & Z
  } \]
  is the pullback of
  $ \xymatrix{
    X \ar[r]^s & Z & Y\ar[l]_t \\
  } $
  and $Z$ is a terminal object then $P$ is the product of $X$ and $Y$.
\end{example}

\begin{example}[Pullbacks in sets]
  Let $X, Y$ be sets and $s \from X \to Z, t \from Y \to Z$ be functions.
  Then $P := \setMap{\pair{x}{y}}{s(x) = t(y)}$ with the projection functions
  $p_1(x,y) := x$ and $p_2(x, y) = y$ is the pullback of
  $ \xymatrix{
    X \ar[r]^s & Z & Y\ar[l]_t \\
  } $.
  The pullback square commutes by construction.
  Now let $A$ be a set and $f_1 \from A \to X, f_2 \from A \to Y$ be functions with $s \of f_1 = t \of f_2$.
  Then we define $\fbar \from A \to P$ by
  $\fbar(a) := \pair{f_1(a)}{f_2(a)}$.
  This is well defined because for $a \in A$ we have $\fbar(a) = \pair{f_1(a)}{f_2(a)} \in P \Leftrightarrow s(f_1(a) = t(f_2(a))$ which is true by assumption.
  We also have $p_1(\fbar(a)) = f_1(a)$ and $p_2(\fbar(a)) = f_2(a)$ by constructions.
  We also see that $\fbar$ is unique with this property.
\end{example}

\begin{definition}[Pushout]
  \label{def:pushout}
  Let $\cat{A}$ be a category and
  $ \xymatrix{
    X & Z \ar[l]_s \ar[r]^t & Y \\
  } $
  a diagram in $\cat{A}$.
  A \demph{pushout} of this diagram is a \demph{pushout square} \\
  \[ \xymatrix{
    Z \ar[r]_s \ar[d]^t & X \ar[d]_{p_1} \\
    Y \ar[r]^{p_2} & P
  } \]
  with the property that for all other pushout squares
  \[ \xymatrix{
    Z \ar[r]_s \ar[d]^t & X \ar[d]_{f_1} \\
    Y \ar[r]^{f_2} & A
  } \]
  in $\cat{A}$, there exists a unique map $\fbar\from P \to A$ such that
  \[ \xymatrix{
    Z \ar[r]_s \ar[d]^t & X \ar[d]_{p_1} \ar[ddr]^{f_1} & \\
    Y \ar[r]^{p_2} \ar[drr]_{f_2}& P \ar@{.>}[dr]|{\fbar} & \\
    & & A  \\
  } \]
  commutes.
\end{definition}

@@

\section*{Conclusion}
We have seen much analogous definitions and proofs.
Our definitions follow a simple ``pattern'':
\todo{Pattern könnte besser erklärt werden aber nicht zu formal!}
\begin{itemize}
  \item First we have a some objects or a commuting diagram ($D_0$).
  \item Then we ``insert'' a object $P$ and get a new commuting diagram ($D_1$) where we ``connect'' $P$ with the objects of $D_0$ with ``projection functions''.
  \item Then we ``replace'' this object with a new object, called $A$, and replace the ``projection functions'' with new functions. We get a new diagram ($D_2$).
  \item For each $A$ there exists a unique map $\fbar$ such if we ``merge'' $D_1$ and $D_2$ we get a new commuting diagram $D_3$, where $A$ and $P$ are connected with $\fbar$.
\end{itemize}

>>>>

  Let $A$ be another set and $f_1 \from A \to X, f_2 \from A \to Y$ functions.
  If we want to define a function $\fbar \from A \to X \times Y$, such that
  $p_1(\fbar(a)) = f_1(a)$ and $p_2(\fbar(a)) = f_2(a)$ for every $a \in A$ holds,
  there is only one possible way to define $\fbar \from A \to X \times Y$:
  $$\fbar(a) := \pair{f_1(a)}{f_2(a)}.$$
  By construction we have $p_1(\fbar(a)) = p_1(f_1(a), f_2(a)) = f_1(a)$ and $p_2(\fbar(a)) = p_2(f_1(a), f_2(a)) = f_2(a)$.
  If we have another $\ftilde \from A \to X \times Y$ with $p_1(\ftilde(a)) = f_1(a)$ and $p_2(\ftilde(a)) = f_2(a)$ we show $\fbar(a) = \ftilde(a)$ for every $a \in A$:
  $$\fbar(a) = \pair{f_1(a)}{f_2(a)} = \pair{p_1(\ftilde(a)}{p_2(\ftilde(a))} = \ftilde(a).$$
  The last step is because we have $\pair{p_1(z)}{p_2(z)} = z$ for every $z \in X \times Y$.
\end{example}

@@

\begin{proof}
  Let $X, Y \in \cat{A}$ and $P$ be a product of $X$ and $Y$ together with $p_1 \from P \to X, p_2 \from P \to Y$
  and $P'$ be another product of $X$ and $Y$ together with $p'_1 \from P' \to X, p'_2 \from P' \to Y$.
  We apply definition \ref{def:prod} multiple times and get the following commuting diagram:
  \[ \xymatrix{
    & P \ar@/_/[ldd]_{p_1} \ar@/^/[rdd]^{p_2} & \\
    & P \ar[ld]^{p_1} \ar@/_/@{.>}[dd]|{\fbar} \ar@{.>}[u]|{f} \ar[rd]_{p_2} & \\
    X & & Y \\
    & P' \ar[lu]_{p_1'} \ar@/_/@{.>}[uu]|{\ftilde} \ar@{.>}[d]|{f'} \ar[ru]^{p'_2} & \\
    & P' \ar@/^/[luu]^{p'_1} \ar@/_/[ruu]_{p'_2} &
  } \]
  for exactly one $f, f', \fbar, \ftilde$. With the inner triangles we get:
    \begin{alignat*}{2}
    p_1  \of \ftilde &= p'_1 &\qquad p_2  \of \ftilde &= p'_2 \\
    p'_1 \of \fbar   &= p_1  &       p'_2 \of \fbar   &= p_2
  \end{alignat*}
  By substitution we get:
  \begin{alignat*}{2}
    p_1   \of (\ftilde \of \fbar) &= p_1  &\qquad p_2  \of (\ftilde \of \fbar) &= p_2 \\
    p'_1  \of (\fbar \of \ftilde) &= p'_1 &       p'_2 \of (\fbar \of \ftilde) &= p'_2
  \end{alignat*}
  With the outer triangles we get:
  \begin{alignat*}{5}
    p_1  &\;\of&&\; f  &= p_1  &\qquad p_2  &\;\of&\; f  &= p_2 \\
    p'_1 &\;\of&&\; f' &= p'_1 &\qquad p'_2 &\;\of&\; f' &= p'_2
  \end{alignat*}
  Note that $f$ and $f'$ are \emph{unique} and the first equations hold for $f = 1_P$ and $f = \ftilde \of \fbar$, thus $\ftilde \of \fbar = 1_P$.
  The second equations holds for $f' = 1_{P'}$ and for $f' = \fbar \of \ftilde$, thus $\fbar \of \ftilde = 1_{P'}$.
  Thus $P$ and $P'$ are isomorphic. \qedhere
\end{proof}

@@

\begin{example}[Posets]
  \label{ex:prod:poset}
  Let $\pair{A}{\le}$ be a poset.
  A \demph{lower bound} for two objects $x, y \in A$ is a element $a \in A$ such that $a \le x$ and $a \le y$.
  A \demph{greatest lower bound} of $x, y \in A$ is a lower bound $z \in A$ of $x$ and $y$ such that for all lower bounds $a \in A$ of $x$ and $y$, $a \le z$.
  With other words: $z$ is a greatest lower bound of $x$ and $y$ if and only if
  $$z \le x \land z \le y \land \forall a \in A, a \le x \land a \le y \Rightarrow a \le z.$$

  When we see the poset $\pair{A}{\le}$ as a category $\catA$ and have a greatest lower bound $z$ of $x, y \in A$, then $z$ is the product of $x$ and $y$.
  To see this, we apply the definition \ref{def:prod}.
  Let $x, y \in \catA$ and $z \in \catA$ be a greatest lower bound of $x$ and $y$.
  Let $a \in \catA$ be another object.
  If we have $f_1 \from a \to x$ and $f_2 \from a \to y$ then by definition of $\catA$, $a \le x$ and $a \le y$ and thus $a$ is a lower bound of $x$ and $y$.
  Now because $z$ is a greatest lower bound, $a \le z$ holds and therefore there is a unique map $\fbar \from a \to z$.
  The diagram commutes trivially.
\end{example}

\begin{example}[Direct sum of vector spaces]
  \label{ex:prod:vspace}
  Let $K$ be a field with the operations $+_K$ and $\cdot_K$ and $V, W$ vector spaces over this field with the operations $+_V, \cdot_V, +_W, \cdot_W$.
  Then we define\footnote{One should verify that the defined structure is indeed a vector field.}
  the \demph{direct sum} $V \oplus W := \bigl(V \times W, +, \cdot \bigr)$ as a new vector field with the following operations:
  \begin{alignat*}{4}
    \pair{v_1}{w_1} + \pair{v_2}{w_2} &:= \pair{v_1 +_V v_2}{w_1 +_W w_2}           &\quad& \forall v_1, v_2 \in V, \forall w_1, w_2 \in W \\
    \alpha \cdot \pair{v}{w}          &:= \pair{\alpha \cdot_V v}{\alpha \cdot_W w} &     & \forall v \in V, \forall w \in W
  \end{alignat*}
  We can show that $V \oplus W$ is the product of $V$ and $W$ in $\Vect$.
  The proof follows the same structure as in example \ref{ex:prod:cart}.
\end{example}

\todo{Soll ich die Verallgemeinerten Produkte über Klassen von Objekten auch einführen?}

@@

\begin{example}[Direct sum of vector spaces]
  \label{ex:coprod:vspace}
  In the example \ref{ex:prod:vspace} we have seen that the direct sum of two vector spaces $V$ and $W$ is the product of $V$ and $W$ in the category $\Vect$.
  We will now see that it is \emph{also} the coproduct.
  In general products and coproducts are not the same as we have seen in the examples \ref{ex:prod:cart} and \ref{ex:coprod:dunion}.
  First we have to define the two coprojection functions $p_1 \from V \to V \oplus W, p_2 \from W \to V \oplus W$:
  \begin{alignat*}{2}
    p_1(x) &:= \pair{x}{0} \\
    p_2(y) &:= \pair{0}{y}
  \end{alignat*}
  Then let $A \in \Vect$ and $f_1 \from V \to A, f_2 \from W \to A$ be vector morphisms.
  We define the vector morphism $\fbar \from V \oplus W \to A$ with $\fbar(v, w) := f_1(v) + f_2(w)$.
  By construction and because $f_1$ and $f_2$ are linear maps we have:
  \begin{alignat*}{4}
    \fbar(p_1(v)) &= \fbar(v, 0) &= f_1(v) + f_2(0) &= f_1(v) \\
    \fbar(p_2(w)) &= \fbar(0, w) &= f_1(0) + f_2(w) &= f_2(w).
  \end{alignat*}
  Now we have to show that $\fbar$ is the only vector morphism with this property.
  Let $\ftilde \from V \oplus W \to A$ be another vector morphism with $\ftilde(p_1(v)) = f_1(v)$ and $\ftilde(p_2(w)) = f_2(w)$.
  Then we have for every $v \in V$ and $w \in W$:
  $$ \fbar(v,w) = f_1(v) + f_2(w)\ = \ftilde(p_1(v)) + \ftilde(p_2(w)) = \ftilde(p_1(v) + p_2(w)) = \ftilde(v, w). $$
  Thus $\fbar = \ftilde$ and $V \oplus W$ is indeed the product of $V$ and $W$ in the category $\Vect$.
\end{example}

\section*{Equalizers and Coequalizers}

@@

\begin{definition}[Coequalizer]
  \label{def:coequa}
  Let $\cat{A}$ be a category and
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y
  } $
  a diagram in $\cat{A}$.
  A \demph{coequalizer} of this diagram is a commuting diagram called \demph{cofork}
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^i & C
  } $
  with the property that for all other coforks
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^f & A
  } $
  in $\cat{A}$, there exists a unique map $\fbar\from C \to A$ such that
  \[ \xymatrix{
    & & A  \\
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[ru]^f \ar[rd]_i \\
    & & C \ar@{.>}[uu]|{\fbar}
  } \]
  commutes.
\end{definition}

@@

\begin{definition}[Symmetric reflexive transitive closure]
  Let $R$ be a binary relation over a set $X$.
  $\hat{R} := R \cup R^{-1}$ is the symmetric closure of $R$ where $R^{-1}$ is defined by
  $R^{-1} := \setMap{\pair{b}{a}}{\pair{a}{b} \in R}$.
  The \demph{symmetric transitive closure} of $R$ is $\sim := \bigcup_{n\in\nat} {\hat R}^n$
  where ${\hat R}^n$ is recursively defined with
  ${\hat R}^0 := \setMap{\pair{a}{a}}{a \in X}$ and
  ${\hat R}^{n+1} := {\hat R}^n \of \hat R = \setMap{\pair{a}{c}}{\pair{a}{b} \in {\hat R}^n \land \pair{b}{c} \in \hat R}$.
  It can be shown that $\sim$ is the \emph{least equivalence relation} containing $R$.
\end{definition}

@@

  To define the function $\fbar \from C \to A$ we need a choice function $\choice \cdot \from C \to Y$ that returns for every equivalence class of $\sim$ a member of it, i.~e.
  $\forall c \in C, \forall y \in Y, y \in c \Leftrightarrow y \sim \choice{c}$
  which is equivalent to $\forall c \in C, \class {\choice c} = c$.

@@

\begin{definition}[Pullback]
  \label{def:pullback}
  Let $\cat{A}$ be a category and
  $ \xymatrix{
    X \ar[r]^s & Z & Y\ar[l]_t \\
  } $
  a diagram in $\cat{A}$.
  A \demph{pullback} of this diagram is a diagram called \demph{pullback square}
  \[ \xymatrix{
    P \ar[r]_{p_2} \ar[d]^{p_1} & Y \ar[d]_{t} \\
    X \ar[r]^s & Z
  } \]
  with the property that for all other pull back squares
  \[ \xymatrix{
    A \ar[r]_{f_2} \ar[d]^{f_1} & Y \ar[d]_{t} \\
    X \ar[r]^s & Z
  } \]
  in $\cat{A}$, there exists a unique map $\fbar\from A \to P$ such that
  \[ \xymatrix{
    A \ar[drr]^{f_2} \ar[ddr]_{f_1} \ar@{.>}[dr]|{\fbar} & & \\
    & P \ar[r]_{p_2} \ar[d]^{p_1} & Y \ar[d]_{t} \\
    & X \ar[r]^s & Z
  } \]
  commutes.
\end{definition}

\begin{example}[Pullbacks and products]
  When the pullback square
  \[ \xymatrix{
    P \ar[r]_{p_2} \ar[d]^{p_1} & Y \ar[d]_{t} \\
    X \ar[r]^s & Z
  } \]
  is the pullback of
  $ \xymatrix{
    X \ar[r]^s & Z & Y\ar[l]_t \\
  } $
  and $Z$ is a terminal object then $P$ is the product of $X$ and $Y$.
\end{example}

\begin{example}[Pullbacks in sets]
  Let $X, Y, Z$ be sets and $s \from X \to Z, t \from Y \to Z$ be functions.
  Then $P := \setMap{\pair{x}{y}}{s(x) = t(y)}$ with the projection functions
  $p_1(x,y) := x$ and $p_2(x, y) := y$ is the pullback of
  $ \xymatrix{
    X \ar[r]^s & Z & Y\ar[l]_t \\
  } $.
  The pullback square commutes by construction.
  Now let $A$ be a set and $f_1 \from A \to X, f_2 \from A \to Y$ be functions with $s \of f_1 = t \of f_2$.
  Then we define $\fbar \from A \to P$ by
  $\fbar(a) := \pair{f_1(a)}{f_2(a)}$.
  This is well defined because for $a \in A$ we have $\fbar(a) = \pair{f_1(a)}{f_2(a)} \in P \Leftrightarrow s(f_1(a)) = t(f_2(a))$ which is true by assumption.
  We also have $p_1(\fbar(a)) = f_1(a)$ and $p_2(\fbar(a)) = f_2(a)$ by constructions.
  We also see that $\fbar$ is unique in respect to that property.
\end{example}

\begin{definition}[Pushout]
  \label{def:pushout}
  Let $\cat{A}$ be a category and
  $ \xymatrix{
    X & Z \ar[l]_s \ar[r]^t & Y \\
  } $
  a diagram in $\cat{A}$.
  A \demph{pushout} of this diagram is a commuting diagram called \demph{pushout square} \\
  \[ \xymatrix{
    Z \ar[r]_s \ar[d]^t & X \ar[d]_{p_1} \\
    Y \ar[r]^{p_2} & P
  } \]
  with the property that for all other pushout squares
  \[ \xymatrix{
    Z \ar[r]_s \ar[d]^t & X \ar[d]_{f_1} \\
    Y \ar[r]^{f_2} & A
  } \]
  in $\cat{A}$, there exists a unique map $\fbar\from P \to A$ such that
  \[ \xymatrix{
    Z \ar[r]_s \ar[d]^t & X \ar[d]_{p_1} \ar[ddr]^{f_1} & \\
    Y \ar[r]^{p_2} \ar[drr]_{f_2}& P \ar@{.>}[dr]|{\fbar} & \\
    & & A  \\
  } \]
  commutes.
\end{definition}

@@

\section*{Conclusion}
We have seen much analogous definitions.
Our definitions follow a simple ``pattern'':
\todo{Pattern könnte besser erklärt werden aber nicht zu formal!}
\begin{itemize}
  \item First we have some objects or a commuting diagram ($D_0$).
  \item Then we ``insert'' an object $P$ and get a new commuting diagram ($D_1$) where we ``connect'' $P$ with the objects of $D_0$ with ``projection functions''.
  \item Then we ``replace'' this object with a new object, called $A$, and replace the ``projection functions'' with new functions. We get a new diagram ($D_2$).
  \item For each $A$ there exists a unique map $\fbar$ such if we ``merge'' $D_1$ and $D_2$ we get a new commuting diagram $D_3$, where $A$ and $P$ are connected with $\fbar$.
\end{itemize}
